Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-08 manuscript/manuscript.tex
edits and start of M&M

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) in some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and normalize the entries for
heterozygosity (see Supplementary Material Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction} for details on
the bias correction). The covariances are expected to be zero when only drift
is acting, as only heritable variation for fitness can create covariance
between allele frequency changes in a closed population
\parencite{Buffalo2019-io}.  Averaging across the ten replicate temporal
covariances matrices, we find temporal covariances that are statistically
significant (95\% block bootstraps CIs do not contain zero), consistent with
linked selection perturbing genome-wide allele frequency changes over very
short time periods. The covariances between all adjacent time intervals are
positive and then decay towards zero as we look at more distant time intervals,
as expected when directional selection affects linked variants' frequency
trajectories until ultimately linkage disequilibrium and the additive genetic
variance for fitness associated with neutral alleles decays
\parencite{Buffalo2019-io}. The temporal covariances per replicate are noisier
but this general pattern holds; see Supplementary Figures
\ref{suppfig:barghi-cov-panels}.  \textcite{Barghi2019-qy}'s design means that
the covariances we see in adjacent time intervals are on average ten
generations apart, and given the temporal decay in covariance we see, the
covariances on shorter time-scales (e.g. if adjacent generations had been
sequenced) may well be higher yet (see Supplementary Material Section
\ref{supp:barghi-covs} for more details).

@@

\vince{I think the bulk of this can go to what will become Materials and Methods?}

The presence of an unselected control line provides an alternative way to
partition the effects of linked selection and genetic drift: we can compare the
total variance in allele frequency change of the control line (which excludes
the effect of artificial selection on allele frequencies) to the total variance
in frequency change of the Longshanks selection lines.  We can partition the
allele frequency change between the two timepoints (20 generations) for a
Longshanks line as \vb{$\Delta p_{t,\mathrm{LS1}} = \Delta_{_{D}}
p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} + \Delta_{_S}
p_{t,\mathrm{LS}}$} where these  terms in are the decomposition are the allele
frequency change due to drift in \vb{Longshanks replicate 1 ($\Delta_{_D}
p_{t,\mathrm{LS1}}$), selection unique to the LS1 replicate ($\Delta_{_U}
p_{t,\mathrm{LS1}}$), and selection response shared between the two Longshanks
replicates ($\Delta_{_S} p_{t,\mathrm{LS}}$) respectively (and similarly for
the Longshanks two line, LS2)}. By construction we will assume that each of
these terms are uncorrelated within replicates, and that only the shared term
covaries between the replicates. Assuming that we can approximate the
contribution of genetic drift in the LS lines the variance in allele frequency
change observed in the control, i.e. \vb{$\var(\Delta p_{t,\mathrm{Ctrl}}) =
\var(\Delta_{_{D}} p_{t,\mathrm{LS2}})=\var(\Delta_{_{D}}
p_{t,\mathrm{LS2}})$}, then we can estimate the increase in variance in allele
frequency change due to selection as $(\var(\Delta p_{t,\mathrm{LS1}}) +
\var(\Delta p_{t,\mathrm{LS2}}))/2 - \var(\Delta p_{t,\mathrm{Ctrl}})$ and the
shared effect of selection across selected lines as $\cov(\Delta
p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}})$.  We estimate at least 32\%
(95\% CI $[21\%, 48\%]$) of the variance in allele frequency change is driven
by the effects of selection, of which 14\% (95\% CI $[3\%, 33\%]$) is estimated
to be unique to a selection line, and 17\% (95\% CI $[9\%, 23\%]$) is the
effect of shared selection between the two Longshanks selection lines.

@@

@@

\begin{table}
  \centering
  \footnotesize
  \centering
  \begin{tabular}{l c c c c c c}
    s & t & 5\% CI & median & 95\% CI & trimmed mean \\ \hline

     0 & 10 & 15.3 & 16.3 & 17.5 & 0.19 \\ \hline
     0 & 20 & 2.8 & 3.7 & 4.6 & 0.05 \\ \hline
     0 & 30 & 4.0 & 4.8 & 5.9 & 0.05 \\ \hline
     0 & 40 & -0.1 & 0.6 & 1.5 & 0.0 \\ \hline
     0 & 50 & -2.7 & -2.0 & -1.3 & -0.03 \\ \hline
     10 & 20 & 14.3 & 15.5 & 16.6 & 0.17 \\ \hline
     10 & 30 & 3.4 & 4.4 & 5.4 & 0.05 \\ \hline
     10 & 40 & 1.5 & 2.3 & 3.3 & 0.03 \\ \hline
     10 & 50 & -4.4 & -3.5 & -2.9 & -0.03 \\ \hline
     20 & 30 & 18.6 & 19.8 & 21.1 & 0.22 \\ \hline
     20 & 40 & 7.0 & 7.9 & 8.9 & 0.09 \\ \hline
     20 & 50 & 0.4 & 1.2 & 2.1 & 0.02 \\ \hline
     30 & 40 & 12.1 & 13.0 & 14.3 & 0.14 \\ \hline
     30 & 50 & -0.4 & 0.7 & 1.8 & 0.01 \\ \hline
     40 & 50 & 12.7 & 13.6 & 14.5 & 0.15 \\ \hline

>>>>

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) in some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and normalize the entries for
heterozygosity. The covariances are expected to be zero when only drift is
acting, as only heritable variation for fitness can create covariance between
allele frequency changes in a closed population \parencite{Buffalo2019-io}.
Averaging across the ten replicate temporal covariances matrices, we find
temporal covariances that are statistically significant (95\% block bootstraps
CIs do not contain zero), consistent with linked selection perturbing
genome-wide allele frequency changes over very short time periods. The
covariances between all adjacent time intervals are positive and then decay
towards zero as we look at more distant time intervals, as expected when
directional selection affects linked variants' frequency trajectories until
ultimately linkage disequilibrium and the additive genetic variance for fitness
associated with neutral alleles decays \parencite{Buffalo2019-io}. The temporal
covariances per replicate are noisier but this general pattern holds; see
Supplementary Figures \ref{suppfig:barghi-cov-panels}.
\textcite{Barghi2019-qy}'s design means that the covariances we see in adjacent
time intervals are on average ten generations apart, and given the temporal
decay in covariance we see, the covariances on shorter time-scales (e.g. if
adjacent generations had been sequenced) may well be higher yet (see
Supplementary Material Section \ref{supp:barghi-covs} for more details).

@@

\vb{The presence of an unselected control line provides an alternative way to
partition the effects of linked selection and genetic drift: we can compare the
total variance in allele frequency change of the control line (which excludes
the effect of artificial selection on allele frequencies) to the total variance
in frequency change of the Longshanks selection lines. We estimate at least
32\% (95\% CI $[21\%, 48\%]$) of the variance in allele frequency change is
driven by the effects of selection, of which 14\% (95\% CI $[3\%, 33\%]$) is
estimated to be unique to a selection line, and 17\% (95\% CI $[9\%, 23\%]$) is
the effect of shared selection between the two Longshanks selection lines.}

@@

\Section{Materials and Methods}

\subsection{Datasets Analyzed}

\subsection{Partitioning Unique and Shared Selection Effects}

We can partition the allele frequency change between the two timepoints (20
generations) for a Longshanks line as \vb{$\Delta p_{t,\mathrm{LS1}} =
\Delta_{_{D}} p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} +
\Delta_{_S} p_{t,\mathrm{LS}}$} where these  terms in are the decomposition are
the allele frequency change due to drift in \vb{Longshanks replicate 1
  ($\Delta_{_D} p_{t,\mathrm{LS1}}$), selection unique to the LS1 replicate
  ($\Delta_{_U} p_{t,\mathrm{LS1}}$), and selection response shared between the
two Longshanks replicates ($\Delta_{_S} p_{t,\mathrm{LS}}$) respectively (and
similarly for the Longshanks two line, LS2)}. By construction we will assume
that each of these terms are uncorrelated within replicates, and that only the
shared term covaries between the replicates. Assuming that we can approximate
the contribution of genetic drift in the LS lines the variance in allele
frequency change observed in the control, i.e. \vb{$\var(\Delta
p_{t,\mathrm{Ctrl}}) = \var(\Delta_{_{D}}
p_{t,\mathrm{LS2}})=\var(\Delta_{_{D}} p_{t,\mathrm{LS2}})$}, then we can
estimate the increase in variance in allele frequency change due to selection
as $(\var(\Delta p_{t,\mathrm{LS1}}) + \var(\Delta p_{t,\mathrm{LS2}}))/2 -
\var(\Delta p_{t,\mathrm{Ctrl}})$ and the shared effect of selection across
selected lines as $\cov(\Delta p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}})$.

\subsection{Variance and Covariance Estimates}

We calculate the variance-covariance matrix of allele frequency changes using a
Python software package we have written, available at
\url{http://github.com/vsbuffalo/cvtk}. Our method uses the sampling depth and
number of diploid individuals sampled to correct the bias in the variance
estimates and a bias that occurs in covariance estimates between adjacent
timepoints (see Supplementary Material Sections \ref{supp:ind-depth-var-corr},
\ref{supp:cov-corr}, and \ref{supp:matrix-correction} for mathematical details
of these estimators).

\subsection{Block Bootstrap}

Throughout the paper, we make use of a block-bootstrapping procedure to

@@

\begin{table}
  \centering
  \footnotesize
  \centering
  \begin{tabular}{l c c c c c c}

    s & t & median & median 95\% CI & trimmed mean & trimmed mean 95\% CI \\ \hline
    0 & 10 & 1.629 & $[1.532, 1.738]$ & 1.874 & $[1.777, 1.969]$ \\
    0 & 20 & 0.371 & $[0.276, 0.465]$ & 0.491 & $[0.403, 0.585]$ \\
    0 & 30 & 0.479 & $[0.4, 0.589]$ & 0.516 & $[0.434, 0.602]$ \\
    0 & 40 & 0.059 & $[-0.012, 0.15]$ & 0.027 & $[-0.05, 0.099]$ \\
    0 & 50 & -0.204 & $[-0.271, -0.125]$ & -0.259 & $[-0.329, -0.187]$ \\
    10 & 20 & 1.549 & $[1.427, 1.659]$ & 1.722 & $[1.617, 1.83]$ \\
    10 & 30 & 0.438 & $[0.339, 0.539]$ & 0.506 & $[0.399, 0.609]$ \\
    10 & 40 & 0.233 & $[0.149, 0.328]$ & 0.254 & $[0.159, 0.343]$ \\
    10 & 50 & -0.355 & $[-0.454, -0.289]$ & -0.319 & $[-0.401, -0.237]$ \\
    20 & 30 & 1.981 & $[1.856, 2.095]$ & 2.195 & $[2.084, 2.302]$ \\
    20 & 40 & 0.792 & $[0.698, 0.894]$ & 0.903 & $[0.815, 0.999]$ \\
    20 & 50 & 0.123 & $[0.042, 0.207]$ & 0.221 & $[0.141, 0.309]$ \\
    30 & 40 & 1.296 & $[1.208, 1.425]$ & 1.385 & $[1.287, 1.483]$ \\
    30 & 50 & 0.07 & $[-0.037, 0.183]$ & 0.116 & $[0.023, 0.21]$ \\
    40 & 50 & 1.36 & $[1.271, 1.446]$ & 1.513 & $[1.427, 1.601]$ \\
